Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-10 pivot/paper.tex
Filtering section

\begin{figure}[t]
\begin{equation*}
\begin{array}{ll}
  \ident{pivot}(F, D) = C, \{ C \mapsto (l, L_1 \cup L_2 \cup \ldots  \}\quad \ball{1}\\[0.5em]
  ~ l = \kvd{type}~C(x:\ident{Query}) = \\
  \quad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F, D)\\
  \quad \qquad \kvd{member}~\qident{filter data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{filter}(F, D)\\
  \quad \qquad (\ldots)\\
\\
\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{2}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{3}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{4} \\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
\quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{5}
\end{array}
\end{equation*}

@@

Add $D$ representing input data (in reality, we do not need all of it; just a function)

zz

>>>>

\begin{figure}[t]
\begin{equation*}
\begin{array}{ll}
  \ident{pivot}(F, D) = C, \{ C \mapsto (l, L_1 \cup L_2 \cup \ldots  \}\quad \ball{1}\\[0.5em]
  ~ l = \kvd{type}~C(x:\ident{Query}) = \\
  \quad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F, D)\\
  \quad \qquad \kvd{member}~\qident{filter data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{filter}(F, D)\\
  \quad \qquad (\ldots)\\
\\
\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\quad\ball{2}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{3}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{4}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{5} \\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
\quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{6}
\end{array}
\end{equation*}

\caption{Pivot type provider -- grouping and aggregation}
\label{fig:tp-filter}
\end{figure}

@@

The example given in Section~\ref{sec:intro} obtained top 8 athletes based on the number
of gold medals from Rio 2016. It used two operations that were omitted in the formalisation in
Section~\ref{sec:pivot-core}. We omitted paging to keep the host language simple, but we also
omitted filtering, which lets us write $\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}$.
This operation is worth further discussion. To support it, the type provider needs not only the
schema of the data set, but also sample data set that is used to offer the available values such 
as $\qident{Rio (2016)}$.

\noindent
In the revised model, the \ident{pivot} function which models the type provider takes the schema
$F$ together with sample data $D$ and provides the type with class context:
%
\begin{equation*}
\begin{array}{l}
\ident{pivot}(F, D) = C, L \quad \textnormal{where}\\
\qquad F=\{ f_1 \mapsto \tau_1, \ldots, f_n \mapsto \tau_n \}\\
\qquad D=\{ f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} 
\end{array}
\end{equation*}

\noindent
In prior work on type providers [X], the actual input value is not available when writing the code
and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
data set is often available at the time of writing the code and so $D$ can be the actual data set.
However, it could equally be a sample or representative subset of the data.

The Figure~\ref{fig:tp-filter} shows a revised version of the \ident{pivot} function \ball{1} 
together with one of the operations discussed before and the newly added \ident{filter} function.

As members members performing data transformations are generated, the provider applies the same
transformation on the sample data. For example, the revised \ident{drop} function \ball{2} takes
the sample data set $D$; when calling \ident{drop} recursively to generate nested class after
dropping a column \ball{3}, it removes the column from the schema (as before), but it also removes
the column from the sample dataset. This means that as nested types are provided, the sample data
used is always representative of data what will be passed to the class at runtime.

After choosing the \qident{filter data} member, the class provided by \ident{filter} lets the 
user select one of the columns \ball{4} based on the schema. The \ident{filter-val} function then 
generates a class with members based on the available values for the specified column in the data 
$D$ \ball{5}. The predicate that filters data based on the value \ball{6} is used both in the 
provided runtime code and also when restricting the sample data set using $\sigma_{\varphi_v}(D)$ 
before providing further filtering by recursively calling \ident{filter}.

The fact that we transform the sample data when providing types is important for two reasons.
It makes it possible to apply filtering after aggregation (which changes the format of data)
and it means that more appropriate values are provided for faceted data. For example,
Figure~\ref{fig:pivot-filter} shows some of the provided members when filtering data by medal,
team and individual athlete. Once we refine the team using $\qident{Team is}.\ident{Mongolia}$
and attempt to filter by athlete using \qident{Athlete is}, the type provider offers only the
names of Mongolian athletes.

\begin{figure}
\begin{equation*}
\hspace{-1em}
\begin{array}{l}
\ident{olympics}.\qident{filter data}.\qident{Medal is}.\ident{Gold}.\qident{Team is}\\[0.25em]
\quad\begin{array}{l}
\rightarrow \qident{Czech Republic}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Barbora Spotakova}\\
\qquad \rightarrow \qident{David Kostelecky}\\
\qquad \rightarrow \qident{David Svoboda}\\
\end{array}\qquad
\begin{array}{l}
\rightarrow \ident{Mongolia}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Badar-Uugan Enkhbat}\\
\qquad \rightarrow \qident{Tuvshinbayar Naidan}\\
\\
\end{array}
\end{array}
\end{equation*}
\caption{Subset of members provided by the filtering operation}
\label{fig:pivot-filter}
\end{figure}
